---
title: "Weather Effect on Public Health and Economics"
author: "Kapil Wadodkar"
date: "Saturday, October 25, 2014"
output: html_document
---
## Abtract:
This analysis is for impact of Storms and other Weather Events on Human life. The impact is analyzed from perspective of Fatalities, Injuries and Economics. This analysis can lead to precautionary measures including budgetery preparedness.
The data is beig referred from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Steps:
* Download and load data

```{r echo=FALSE, results='hide',message=FALSE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FstormData.csv.bz2", 
              "repdata_data_stormData.csv.bz2")

baseData <- bzfile("repdata_data_stormData.csv.bz2", "repdata_data_stormData.csv")
stormData <- read.csv2(baseData, sep = ",", stringsAsFactors = FALSE)
```

* Load necessary Libraries

```{r echo=FALSE, results='hide',message=FALSE}
library(data.table)
library(ggplot2)
library(scales)
```

* Cleaning and organizing data
```{r}
`echo = FALSE`
str(stormData)
summary(stormData)
head(stormData)
stormData$FATALITIES <- as.numeric(stormData$FATALITIES)
stormData$INJURIES <- as.numeric(stormData$INJURIES)
stormData$EVTYPE <- toupper(stormData$EVTYPE)
eventType <- sort(unique(stormData$EVTYPE))
```
Selectin top 50 records for representaion. 
Collecting top 20 represtntational records for Fatalities vs. Weather Events.
```{r}
eventType[1:50]
stormData$EVTYPE <- as.factor(stormData$EVTYPE)
fatalities <- as.data.table(subset(aggregate(FATALITIES ~ EVTYPE, data = stormData, 
                                             FUN = "sum"), FATALITIES > 0))
fatalities <- fatalities[order(-FATALITIES), ]
top20Fatalities <- fatalities[1:20, ]
```
The first plot for Fatalities vs. Weather Events. (Coordinates flipped in every plot for better visibility of lables)
```{r}
ggplot(data = top20Fatalities, aes(EVTYPE, FATALITIES, fill = FATALITIES)) + geom_bar(stat = "identity") + 
  xlab("Event") + ylab("Fatalities") + coord_flip() + ggtitle("Fatalities caused by Events (top 20) ") + 
  theme(legend.position = "none")
```
Representational data of top 20 records for Injuries vs. Weathre Events. 
```{r}
injuries <- as.data.table(subset(aggregate(INJURIES ~ EVTYPE, data = stormData, 
                                           FUN = "sum"), INJURIES > 0))
injuries <- injuries[order(-INJURIES), ]
top20Injuries <- injuries[1:20, ]
ggplot(data = top20Injuries, aes(EVTYPE, INJURIES, fill = INJURIES)) + geom_bar(stat = "identity") + 
  xlab("Event") + ylab("Injuries") + coord_flip() + ggtitle("Injuries caused by Events (top 20) ") + 
  theme(legend.position = "none")
```
Now while calculating the Cost factor the data is being extrapolated in accordance with the data present as we are operating only on top 20 records as representation.
```{r echo=FALSE, results='hide',message=FALSE}
unique(stormData$PROPDMGEXP)
stormData$PROPDMGEXP <- toupper(stormData$PROPDMGEXP)
unique(stormData$PROPDMGEXP)
table(stormData$PROPDMGEXP)
calcExp <- function(x, exp = "") {
  switch(exp, `-` = x * -1, `?` = x, `+` = x, `1` = x, `2` = x * (10^2), `3` = x * 
           (10^3), `4` = x * (10^4), `5` = x * (10^5), `6` = x * (10^6), `7` = x * 
           (10^7), `8` = x * (10^8), H = x * 100, K = x * 1000, M = x * 1e+06, 
         B = x * 1e+09, x)
}
applyCalcExp <- function(vx, vexp) {
  if (length(vx) != length(vexp)) 
    stop("Not same size")
  result <- rep(0, length(vx))
  for (i in 1:length(vx)) {
    result[i] <- calcExp(vx[i], vexp[i])
  }
  result
}
```
While extrapolating Scales library can be of us.
Considering top 20 records for Econimic Costs vs. Weather Events.
```{r}
stormData$EconomicCosts <- applyCalcExp(as.numeric(stormData$PROPDMG), stormData$PROPDMGEXP)
summary(stormData$EconomicCosts)
costs <- as.data.table(subset(aggregate(EconomicCosts ~ EVTYPE, data = stormData, 
                                        FUN = "sum"), EconomicCosts > 0))
costs <- costs[order(-EconomicCosts), ]
top20Costs <- costs[1:20, ]
ggplot(data = top20Costs, aes(EVTYPE, EconomicCosts, fill = EconomicCosts)) + geom_bar(stat = "identity") + 
  scale_y_continuous(labels = comma) + xlab("Event") + ylab("Economic costs in $") + coord_flip() +
  ggtitle("Economic costs caused by Events (top 20) ") +  theme(legend.position = "none")
```
